Physcraper: A Python package for continually updated gene trees

### 2020-05-11 physcraper_ms.Rmd
methods

---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: 'Physcraper: A Python package for continually updated gene trees'
tags:
  - Python
  - phylogenetics
  - update
  - species tree
  - Open Tree
authors:
  - name: Luna L. Sanchez Reyes
    orcid: 0000-0000-0000-0000
    affiliation: 1 # (Multiple affiliations must be quoted, like this: "1, 2")
  - name: Emily Jane McTavish
    orcid: 0000-0000-0000-0000
    affiliation: 1
affiliations:
 - name: University of California, Merced
   index: 1
 - name: Institution 2
   index: 2
citation_author: Sanchez-Reyes et. al.
date: 24 April 2020
year: 2020
bibliography: paper.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  eval = TRUE
)
```

# Summary

1. Phylogenies are still a key part of research in all areas of biology. The nonspecialist
can use one of the many tools available to automatize phylogenetic reconstruction.
However, interpretation of results, comparison with previously available phylogenetic
hypotheses, and choosing of one phylogeny for discussion still impose difficulties
to one that is not a specialist on phylogenetic methods or on a particular group of study.

1. Physcraper is an open‐source Python program that automatizes the update of public
phylogenies by making use of public DNA sequence data and taxonomic information.

1. Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
hypothesis based on public expert phylogentic knowledge.
The phylogenetics and group specialist will find it useful as a tool to facilitate comparison
of alternative phylogenetic hypotheses (topologies).
***Is physcraper intended for the nonspecialist?? We have two types of nonspecialists:
the ones that do not know about phylogenetic methods and the ones that might know
about phylogenetic methods but do not know of a certain biological group.***

@@

1. We hope the physcraper workflow demonstrates the benefits of open results in phylogenetics and encourages researchers to strive for better data sharing practices.

@@

There is a lot of phylogenetic knowledge already available.
Some of it has been made public at OToL.
Most or all public phylogenies have the potential to be updated with the ever increasing
amount of genetic data that is available on GenBank.
There are now various methods to automatize use of GenBank data for phylogenetic reconstruction, which are
generally intended to facilitate the use of phylogenetic knowledge for the nonspecialist.
However, none of them provide a means to compare with previously available expert
knowledge on a particular group.

A key aspect of the standard phylogenetic workflow is comparison with already existing
phylogenetic hypotheses and with phylogenies that are considered "best" by experts
not only in phylogenetics, but also experts of the focal group of study.

@@

# Physcraper overview

@@

```{r fungi-phylo, eval = TRUE, fig.retina = 2, fig.height = 4}
ape::plot.phylo(my_trees[[1]], type = "fan", cex = 0.1)
```

>>>>

# Abstract

1. Phylogenies are a key part of research in all areas of biology. Tools that automatize
some parts of the process of phylogenetic reconstruction (mainly character matrix construction)
have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
However, interpretation of results, comparison with previously available phylogenetic
hypotheses, and choosing of one phylogeny for downstream analyses and discussion still impose difficulties
to one that is not a specialist either on phylogenetic methods or on a particular group of study.

1. Physcraper is an open‐source, command-line Python program that automatizes the update of published
phylogenies by making use of public DNA sequence data and taxonomic information,
providing a framework for comparison of phylogenies.

1. Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
hypothesis based on already available expert phylogentic knowledge.
Phylogeneticists and group specialists will find it useful as a tool to facilitate comparison
of alternative phylogenetic hypotheses (topologies).
***Is physcraper intended for the nonspecialist?? We have two types of nonspecialists:
the ones that do not know about phylogenetic methods and the ones that might know
about phylogenetic methods but do not know much acbout a certain biological group.***

@@

1. We hope the physcraper workflow demonstrates the benefits of opening results in phylogenetics and encourages researchers to strive for better data sharing practices.

@@

Phylogenies are important.
Generating phylogenies is not easy.
The process of phylogenetic reconstruction implies many steps that can be generalized to the following:
1. Obtention of molecular or morphological character data -- get DNA from some organisms and sequence it, or get it from an online repository, such as GenBank.
1. Assemble a hypothesis of homology -- Create a matrix of your character data, by aligning the sequences, in the case of molecular data.
1. Analyse this hypothesis of homology to infer phylogenetic relationships among the organisms you are studying -- Use different available programs to infer molecular evolution, trees and times of divergence.
1. Discuss the inferred relationships in the context of previous hypothesis, the biology and biogeography of the organisms, etc. -- Answer the question, *is this phylogenetic solution fair/reasonable?*

Each of these steps require different types of specialized training: in the field,
in the lab, in front of a computer, discussions with experts in the methods, and/or in the biological group of study.
All of these steps also require considerable amounts of time for training and implementation.

In the past decade, various studies have developed solutions to automatize the first and second steps, by creating
pipelines that mine already available molecular data from the GenBank repository,
to obtain homologous characters that can be used for phylogenetic reconstruction.
These tools have been presented as aid for the nonspecialist to decrease some of
the difficulties in the generation of phylogenetic knowledge. However, they are
not that often used as so, suggesting that there are still difficulties for the nonspecialist.
The phylogenetic community has some reserves towards these tools, too. Mainly because they sometimes act as a black box.
However, automatizating the assembly of the character data set is a crucial step towards
reproducibility for a task that was otherwise primarily artisanal and hence largely
non-reproducible.

Even if it is hard to obtain phylogenies, we invest copious amounts of time and energy in generating them.
They are crucial to solve problems such as food security, global warming, global health.
There is a lot of phylogenetic knowledge already available in published peer-reviewed studies.
In this sense, the non-specialists (and also the specialist) face a new problem: how do I choose the best phylogeny.

Public phylogenies can be updated with the ever increasing amount of genetic data that is available on GenBank.

A way to automatize the comparison of phylogenetic hypotheses and to allow reproducibility of the last step of the process.

A key aspect of the standard phylogenetic workflow is comparison with already existing
phylogenetic hypotheses and with phylogenies that are considered "best" by experts
not only in phylogenetics, but also experts on the focal group of study.

@@

# How does Physcraper work?

## Study tree and alignment

- The phylogenetic tree has to be in Open Tree of Life. You can choose form a variety of ytrees supporting any node of the Tree of Life.
- The alignment should be a gene alignment that was used to generate the tree. The alignments are usually stored in treebase or dryad.
  If it is stored in Treebase, it can be downloaded by pyscraper. If it is on dryad, it has to be provided by the user.
- Taxon name matching step: check that taxon names on the tips of the trees are in the character matrix and visceversa.
- A csv file with the summary of taxon name matching is produced for the user.
- Unmatched taxon names are dropped from the tree and alignment. ***What is the minimum amount of tips necessary for a physcraper run??***
- A .tre and .aln files are generated for a physcraper run.

## DNA sequence search and cleaning

- The first step is to identify the taxonomic MRCA of all species in the tree. The taxonomic MRCA
  is the closest phylogenetic MRCA that has a taxonomic name assigned in the NCBI database. ***Does
  physcraper take into account the focal clade from Open Tree in any way???***
- The taxonomic MRCA is used to constrain the DNA sequence search within that taxonomic group on the GenBank database.
- The DNA sequence search can be done on a local database that is easily setup by the user, or remotely.
- The BLAST algorithm is used to identify all DNA sequences in the database within
  the taxonomic MRCA that are similar to each sequence in the alignment ***How does the blast is setup? Is it an all-blast-all??***
- Reverse complement sequences are identified and translated.
- A fasta file containing all sequences resulting from the search is generated for the user.

## DNA sequence alignment

- MUSCLE

## Tree reconstruction

- A gene tree is reconstructed for each alignment provided, using ***what software?***

## Tree comparison

- Conflict information can only be generated in the context of the whole Open Tree of
Life. Otherwise, it is not really possible to get conflict data.
***- One way to compare two independent phylogenetic trees is to compare them both to
the synthetic OToL and then measure how well they do agianst each other***

@@

<!-- ```{r fungi-phylo, eval = TRUE, fig.retina = 2, fig.width = 2} -->
<!-- ape::plot.phylo(ape::ladderize(my_trees[[1]]), type = "phylogram", cex = 0.1) -->
<!-- ``` -->
